A Correct Compiler for Continuation Marks

### 2012-09-20 paper.tex
A bit of a dump but new prose and ideas in there.

@@

>>>>

\usepackage{amsmath}
\usepackage{slatex}

\title{A Correct Compiler for Continuation Marks}

\author{Kimball Germane \and Jay McCarthy}
\institute{Brigham Young University}

% Look at the many other papers on c marks to justify them (see mine for
% example). Talk about how your transformation is a compiler and right
% now no language but Racket has C-marks, so we could use this to get
% them in that language. For Racket though, now they don't need to be
% built in to the compiler if we don't want. In addition, it makes it
% easier to do a type system.

\newcommand{\cm}[0]{$\lambda_{cm}$}
\newcommand{\lv}[0]{$\lambda_v$}
\newcommand{\lc}[0]{$\lambda$-calculus}
\newcommand{\wcm}[2]{(\mathrm{wcm}\,#1\,#2)}
\newcommand{\ccm}[0]{(\mathrm{ccm})}
\newcommand{\app}[2]{(#1\,#2)}
\newcommand{\abs}[2]{(\lambda\,(#1)\,#2)}
\newcommand{\hole}[0]{\bullet}
\newcommand{\rr}[0]{\rightarrow}
\newcommand{\lrrs}[0]{\rightarrow^{*}}
\newcommand{\lvrr}[0]{\rightarrow_v}
\newcommand{\lvrrs}[0]{\rightarrow_v^{*}}
\newcommand{\cmrr}[0]{\rightarrow_{cm}}
\newcommand{\cmrrs}[0]{\rightarrow_{cm}^{*}}

@@

\maketitle

\begin{abstract}
Continuation marks, a programming language feature which generalizes stack inspection, do not yet enjoy a transformation to the plain $\lambda$-calculus. We present a CPS-like transformation from the call-by-value $\lambda$-calculus augmented with continuation marks to the pure call-by-value $\lambda$-calculus. We discuss how such a transformation would simplify the construction of compilers which treat continuation marks correctly. We document an iterative, feedback-based approach using Redex. We accompany Finally, we verify the transformation with a meaning-preservation theorem.
\end{abstract}

\section{Introduction}

Language transformation is useful for a variety of purposes. It can be used to give meaning to the source language in terms of the target language [cite something], provide a characterization more amenable to analysis. Language transformation can also be used to \emph{desugar} a language into a simpler core. This 
Transforming has the following benefits:
reduce core code (desugaring, simplifies implementations)
give meaning to constructs (semantically, as well as, say, macros)
local transformations - macros, for example
global transformations - compilers of all kinds, for example

talk about the common practice of analyzing an extension of the lambda calculus, and a subpractice of doing so via a transformation to the core.

While continuation marks have had significant treatment, they do not currently enjoy a transformation to the plain $\lambda$-calculus.

\section{Continuation Marks}

Continuation marks generalize stack inspection, allowing a bevy of features that rely on some form of it to be defined at the language level. This includes algebraic steppers \cite{clements2001modeling}, security policies \cite{clements2004tail}, and aspects \cite{tucker2003pointcuts}.

Few languages offer continuation marks. Their inception lay in Racket \cite{plt-tr1} and they have been added experimentally to \emph{JavaScript} \cite{clements2008implementing}, but their adoption by other languages has been nonexistent.

Racket offers desugaring facilities to the user via a powerful macro system. This requires and is limited to local transformations limited to expression boundaries.

A transformation-based compiler, such as a CPS compiler \cite{appel2007compiling}, effects a global transformation of the program. This effect discards original source or transforms it so radically that it is suitable only for the levels beneath the programmer's regular interaction.

\section{\cm}

Terms $e$ in \cm\ are defined inductively by
\begin{align}
e=\, &\wcm{e}{e}\\
     &\ccm\\
     &\app{e}{e}\\
     &v\\
     &x
\end{align}
where 
\begin{equation}
v=\abs{x}{e}
\end{equation}
describes the values of the \lc.

The sole novelty of \cm\ above the plain \lc\ is the ability to annotate and observe the continuation. Our evaluation model of \cm, introduced by Pettyjohn et al. \cite{pettyjohn2005continuations}, utilizes the current evaluation context for that purpose. The mutually inductive valuation contexts $E$, $F$ are defined by
\begin{align}
E=\, &\wcm{v}{F}\\
     &F\\
F=\, &\app{E}{e}\\
     &\app{v}{E}\\
     &\wcm{E}{e}\\
     &\hole
\end{align}
where $\hole$ denotes a ``hole'' in the evaluation context, the ultimate destination of the evaluation of the expression that previously resided there.

The semantics of \cm\ are
\begin{align}
E[\app{\abs{x}{e}}{v}]  &\rr E[e[x\leftarrow v]]\\
E[\wcm{v}{\wcm{v'}{e}}] &\rr E[\wcm{v'}{e}]\\
E[\wcm{v}{v'}]          &\rr E[v']\\
E[\ccm]                 &\rr E[\chi(E,\abs{x}{\abs{y}{y}})]
\end{align}
where the metafunction $\chi$ is defined by
\begin{align}
\chi(\wcm{v}{F},vs) &= \chi(F,\abs{z}{\app{\app{z}{v}}{vs}})\\
\chi(\app{E}{e},vs) &= \chi(E,vs)\\
\chi(\app{v}{E},vs) &= \chi(E,vs)\\
\chi(\wcm{E}{e},vs) &= \chi(E,vs)\\
\chi(\hole,vs)      &= vs
\end{align}.

\section{Transformation}

\section{Testing}

A pragmatic approach to the discovery of a correct transformation involves consistent feedback and testing to validate candidate transforms. Testing is no substitute for proof, but, as Klein et al. \cite{klein2012run} show, proof is no substitute for testing. Lightweight mechanization is a fruitful middle ground between pencil-and-paper analysis and fully-mechanized formal proof. We use Redex \cite{findler2010redex}, a domain-specific language for exploring language semantics, to provide feedback, thoroughly exercise candidates, and perform exploratory analysis.

\setkeyword{define-language define-extended-language define-metafunction reduction-relation extend-reduction-relation define-metafunction/extension in-hole hole variable-not-otherwise-mentioned}
\setkeyword{wcm ccm}

\setspecialsymbol{lambda}{$\lambda$}
\setspecialsymbol{->}{$\to$}
\setspecialsymbol{-->}{$\rightarrow$}
\setspecialsymbol{betav}{$\beta$v}
\setconstant{error}
\setkeyword{chi}

\setspecialsymbol{lambdav}{$\lambda$v}
\setspecialsymbol{lambdav-rr}{$\lambda$v-rr}
\setspecialsymbol{lambdav-subst}{$\lambda$v-subst}

\begin{schemedisplay}
(define-language lambdav
  (e (e e) x v error)
  (x variable-not-otherwise-mentioned)
  (v (lambda (x) e))
  (E (E e) (v E) hole))
\end{schemedisplay}

\begin{schemedisplay}
(define lambdav-rr
  (reduction-relation lambdav
   (--> (in-hole E ((lambda (x) e) v))
        (in-hole E (lambdav-subst x v e))
        "betav")
   (--> (in-hole E x)
        (in-hole E error)
        "error: unbound identifier")
   (--> (in-hole E (error e))
        (in-hole E error)
        "error in operator")
   (--> (in-hole E (v error))
        (in-hole E error)
        "error in operand")))
\end{schemedisplay}

\begin{schemedisplay}
(define-metafunction lambdav
  lambdav-subst : x v e -> e
  ;; 1. substitute in application
  [(lambdav-subst x_1 v_1 (e_1 e_2))
   ((lambdav-subst x_1 v_1 e_1) (lambdav-subst x_1 v_1 e_2))]
  ;; 2a. substitute in variable (same)
  [(lambdav-subst x_1 v_1 x_1)
   v_1]
  ;; 2b. substitute in variable (different)
  [(lambdav-subst x_1 v_1 x_2)
   x_2]
  ;; 3a. substitute in abstraction (bound)
  [(lambdav-subst x_1 v_1 (lambda (x_1) e_1))
   (lambda (x_1) e_1)]
  ;; 3b. substitute in abstraction (free)
  [(lambdav-subst x_1 v_1 (lambda (x_2) e_1))
   (lambda (x_2) (lambdav-subst x_1 v_1 e_1))]
  ;; 4. substitute in error
  [(lambdav-subst x_1 v_1 error)
   error])
\end{schemedisplay}

\setspecialsymbol{lambdacm}{$\lambda$cm}
\setspecialsymbol{lambdacm-rr}{$\lambda$cm-rr}
\setspecialsymbol{lambdacm-subst}{$\lambda$cm-subst}

Since \cm\ is a superset of \lv, we need only extend the definition of the \lv\ interpreter to accommodate the additions \cm\ brings.

\begin{schemedisplay}
(define-extended-language lambdacm lambdav
  (e .... (wcm e e) (ccm))
  (E (wcm v F) F)
  (F (E e) (v E) (wcm E e) hole))
\end{schemedisplay}

Redex allows us to easily define a proper extension of a language, inheriting anything left unspecified. As similar as the \lv\ interpeter definition is to the \lv\ definition in figure \ref{lv-language-forms}, this \cm\ interpreter definition is to the \cm\ definition in figure \ref{cm-language-forms}.

\begin{schemedisplay}
(define lambdacm-rr
  (extend-reduction-relation lambdav-rr lambdacm
   (--> (in-hole E ((lambda (x) e) v))
        (in-hole E (lambdacm-subst x v e))
        "betav")
   (--> (in-hole E (wcm v_1 (wcm v_2 e)))
        (in-hole E (wcm v_2 e))
        "wcm-collapse")
   (--> (in-hole E (wcm v_1 v_2))
        (in-hole E v_2)
        "wcm")
   (--> (in-hole E (ccm))
        (in-hole E (chi E (lambda (x) (lambda (y) y))))
        "chi")
   (--> (in-hole E (wcm error e))
        (in-hole E error)
        "error in wcm mark expression")
   (--> (in-hole E (wcm v error))
        (in-hole E error)
        "error in wcm body expression")))
\end{schemedisplay}

The first three rules in the reduction relation correspond with the three additional semantic rules found in \ref{cm-language-semantics}. The remaining handle cases introduced by the the new language forms' interaction with \scheme'error'.

\begin{schemedisplay}
(define-metafunction/extension lambdav-subst lambdacm
  lambdacm-subst : x v e -> e
  ;; 1. substitute in wcm form
  [(lambdacm-subst x_1 v_1 (wcm e_1 e_2))
   (wcm (lambdacm-subst x_1 v_1 e_1) (lambdacm-subst x_1 v_1 e_2))]
  ;; 2. substitute in ccm form
  [(lambdacm-subst x_1 v_1 (ccm))
   (ccm)])
\end{schemedisplay}

The \scheme{lambdacm-subst} metafunction is extended to accommodate the additional forms in \cm.

\begin{schemedisplay}
(define-metafunction lambdacm
  chi : E v -> v
  [(chi hole v_ms)      v_ms]
  [(chi (E e) v_ms)     (chi E v_ms)]
  [(chi (v E) v_ms)     (chi E v_ms)]
  [(chi (wcm E e) v_ms) (chi E v_ms)]
  [(chi (wcm v E) v_ms) (chi E (lambda (p) ((p v) v_ms)))])
\end{schemedisplay}

\section{Proof}

\section{Related Work}

\section{Conclusion}

\bibliographystyle{plain}
\bibliography{paper}
